\documentclass[article]{article}
\usepackage[T1]{fontenc} 
\usepackage[latin9]{inputenc} 
\usepackage{amsmath} 
\usepackage{amstext} 
\usepackage{amsthm} 
\usepackage{amssymb}
\usepackage{mathrsfs} 
\usepackage{hyperref}
\usepackage{babel}   

\makeatletter 

\makeatother


\providecommand{\definitionname}{Definition} 
\providecommand{\theoremname}{Theorem}
\providecommand{\Cech}{\v{C}ech }
\providecommand{\Poinacre}{Poincar\'e }
\providecommand{\Kunneth}{K{\"u}nneth }

\begin{document}
\title{Continuity: Outline and Bibliography}
\author{Rich Cochrane}
\maketitle

\section{General Topology}

\begin{itemize}
	\item{Introduction to General Topology}
	\item{Topology, Logic and Stone Duality}
	\item{Topics in General Topology}
	\item{Knot Theory}
\end{itemize}
This is the part I already have the most complete set of notes for, although they need finishing and there's a lot of pedagogical finesse missing. One thing that's missing at present is some traditional topics in elementary topology (compactness, covering spaces, classification of 2-manifolds). 

The downside if this approach is of course that it takes a lot more work to get to those topics; the advantage is the friends we make along the way. That's kind of a theme for a lot of this project, though.

It would be nice to do something with knot theory but probably that doesn't fit here as it's quite a quirky subject; maybe an additional short course makes sense. I think we could cover some interesting stuff without needing algebraic topology or manifolds.

\begin{itemize}
	\item[]{Vickers, \emph{Topology via Logic}}
	\item[]{Davey and Priestley, \textit{Introduction to Lattices and Order}}
	\item[]{Johnstone, \emph{Stone Spaces}}
	\item[]{Picado and Pultr, \emph{Frames and Locales}}
	\item[]{Tennison, \emph{Sheaf Theory} (Ch 1-3)}
\end{itemize}

\section{Algebraic Topology}

\begin{itemize}
	\item{Introduction to Algebraic Topology}
	\item{Homology of Manifolds}
	\item{Homotopy Theory}
\end{itemize}

I already have some detailed notes on the first half or so of the Introduction course, and a good idea of what I want to cover for Homology of Manifolds. Some thought will need to be given to what is essential to include into the introductory course in terms of supporting the courses for which this will be a prerequisite. For example, in Smooth Manifolds we will need ot be able to express the de Rham Cohomology, which may imply a bit more homology than otherwise.

These courses need to develop all the homological algebra they need, which hopefully won't be much; some results and techniques will have to be taken `on trust' until the reader has done the dedicated Homological Algebra course (where I expect to deal with Tor and Ext, which are indispensable for some of the results we would like to cover here).

I'm less clear on what would be needed for homotopy, but I like that Hu (a) casts the whole thing in terms of extension problems and (b) puts fibre bundles centre stage. I've included Lee because he introduces the elementary stuff that Hu breezes through at speed. I know almost none of this -- all the homotopy I know I learned from Lee's book.

\begin{itemize}
	\item[]{Munkres, \textit{Algebraic Topology}}
	\item[]{Rosiak, \textit{Sheaf Theory Through Examples}}
	\item[]{Wallace, \textit{Algebraic Topology} (later chapters)}
	\item[]{Lee, \textit{Introduction to Topological Manifolds}}
	\item[]{Hu, \textit{Homotopy Theory}}
\end{itemize}

\section{Smooth Manifolds}

\begin{itemize}
	\item{Introduction to Smooth Manifolds}
	\item{Differential and Integral Equations}
\end{itemize}

This block is primarily about developing the Cartan calculus on manifolds from scratch using the `locally ringed space' approach. I can't help but think of this as "partially rigid geometry" even though that's not quite right (we don't get to measure much). 

I think of this as the heart of the project so I'd prefer to keep it very focused. I therefore expect this to be one, longer-then-usual course that covers everything needed for subsequent courses where the fancier stuff happens. Perhaps the connection with differential equations could be hived off into a separate course, though, since I don't think it'll be needed later.

The reading here has a lot of overlaps. I'm intending to get most of what I want from Wedhorn Ch 4, 5, and then 8-10, which has the right emphasis but is probably missing rather a lot. Ramanan is a bit more terse than Wedhorn, which may or may not be an advantage -- only his first 100 pages are relevant here.

I want to include lots of very elementary, very concrete calculations, which might benefit from translating Weintraub into Vaisman's language. At the moment I'm considering that to be the heart of this whole block.

Selectively reading parts of Federer may provide a very unified, structural overview of all the machinery -- I expect that might be useful as the subject can seem a bit chaotic. 

I've included Warner as a potentially helpful supplement -- he proves the de Rham theorem using sheaves and his final chapter is on the Hodge theorem, which I'd also like to get to but not sure that will be possible at this stage (maybe there'll be a later course on harmonic analysis in general, though). 

I'm undecided at this point whether to emphasise the SDG `infinitesimals' perspective when first developing derivatives. If I do, I probably won't need any more than what's in the first part of Bell's essay here: \url{https://citeseerx.ist.psu.edu/document?repid=rep1&type=pdf&doi=9100113e2c74c99c5262c6a686a63b166dc9aaa7}. There's much deeper waters here but a lot of it comes down to debates about mathematical foundations that are highly technical and not really what I'm aiming for. It's possible this point of view can be developed in a separate course within this block.

\begin{itemize}
	\item[]{Wedhorn, \textit{Manifolds, Sheaves and Cohomology}}
	\item[]{Ramanan, \textit{Global Calculus}}
	\item[]{Weintraub, \textit{Differential Forms} (very concrete calculation-based exposition)}
	\item[]{Vaisman, \textit{Cohomology and Differential Forms} (may be helpful to supplement Wedhorn)}
	\item[]{Federer, \textit{Geometric Measure Theory} (parts of Ch 1-4)}
	\item[]{Warner, \textit{Foundations of Differentiable Manifolds and Lie Groups}}
\end{itemize}

\section{Lie Groups}

\begin{itemize}
	\item{Introduction to Lie Groups}
\end{itemize}

Think of this as an optional application of Smooth Manifolds, but it's an important one and maybe we fold them together. I'm assuming I'll hive off the real algebra to another course. 

Since Lie groups are the kinds of symmetries that continuous spaces should have, it makes sense that Thurston's eight geometries of 3-manifolds are defined using them. The spaces on which the Lie groups act are fibre bundles so there's a chance this material actually ends up looking rather attractive at this point. If included, this would just be the `lite' version of geometrization, mostly just showing off the resulting spaces. I don't currently have a good source for this, though -- Thurston's books aren't suitable.

Note there is a lot of material that connects this to smooth manifolds in both Lee and Ramanan, which is probably the place to start to make sure this material doesn't become too disconnected from the main trunk, at least in the introductory course. 

I'd like one day to do a `quantum geometry' or `noncommutative geometry' course following, e.g., Sontz \textit{Principal Bundles: The Quantum Case}. However, this world seems to be very different in character and I'm not sure it fits into the overall scope of this project as I currently understand it. The material is interesting, though, and I might come back to it.

\begin{itemize}
	\item[]{Hilgert and Neeb, \textit{Structure and Geometry of Lie Groups}}
	\item[]{Curtis, \textit{Matrix Groups}}
	\item[]{Thurston, \textit{Geometry and Topology of 3-Manifolds}}
	\item Fecko, \textit{Differential Geometry and Lie Groups for Physicists}
\end{itemize}

\section{Differential Topology}

\begin{itemize}
	\item{Introduction to Differential Topology}
\end{itemize}

I'm seeing the introductory course primarily as an application of the Cartan calculus and I'm hoping there's enough rigidity there to do some interesting topology. If not this might need a re-think or to be scaled down, or it might need to come after Riemannian Geometry.

Guillemin and Pollack give a very elementary introduction; some of this will already have been covered. It will probably need to be supplemented with other texts that contain more topics. It's possible that the GTM volume Hirsch, \textit{Differential Topology} would be useful here, although I believe it completely overlaps with the others.

As an application of Morse theory, Nicolaescu includes a discussion of hamiltonian flows on a manifold. This may be good preparation for the (much more difficult) results we would like to be able to get close to in the Riemannian space involving the Ricci flow. It may be that this whole section has the secret agenda of being a preparation for the same ideas reappearing in a `more rigid' setting where proper geometry is finally possible. Note that Jost (in Riemannian Geometry) also has a chapter on Morse theory that looks good.

One result to aim for might be the h-Cobordism Theorem: \url{https://www.math.uchicago.edu/~may/VIGRE/VIGRE2011/REUPapers/MacKieMason.pdf}. This would probably require the homotopy course though so it should be separate from the introductory material.

I have thoughts of two follow-up courses but I'm not sure whether they're relevant or do-able. A course on Vector Bundles, K Theory and Characteristic Classes could be a nice `advanced diff top' option. A course on 4-Manifolds and Kirby Calculus could be a nice topic to show off how different things can be in different dimensions. For the former, Husemoller is a bit less forgiving than Hatcher, probably has fewer good examples but goes deeper; I'm not sure which, if either, would be the most useful yet or what exactly I'm looking for in this area. And of course Milnor, \textit{Characteristic Classes} is a classic.

\begin{itemize}
	\item[]{Guillemin and Pollack, \textit{Differential Topology}}
	\item[]{Nicolaescu, \textit{Invitation to Morse Theory}}
	\item 
\end{itemize}

\section{Riemannian Geometry}

\begin{itemize}
	\item{Introduction to Riemannian Geometry}
	\item{Geometric Applications of Vector Flows}
\end{itemize}

The slogan here is `geometry is the study of a connection on a principal bundle' -- the goal is to understand that statement and believe it because one has seen it in action. Getting to noncommutative geometry (Sontz) would be great and I think it should be possible to find a through-line that gets us there given all the preparation of the previous sections.  Jost covers topics other introductory textbooks skip and seems to overall set out from a useful starting-point; probably his first four chapters are a good place to start.

Note that Ramanan develops the beginnings of Riemannian geometry and it's probably useful to get the basic machinery form that perspective.

I'm less convinced we can do much with the proof of the Poincar\'e conjecture but it's worth a look to see if there's a part of it that makes sense to include. Morgan and Fong's book looks like being the right path to that if such a thing exists. That's what the second course in this list would be if it looks feasible possible.

\begin{itemize}
	\item[]{Berger, \textit{A Panoramic View of Riemannian Geometry}}
	\item[]{Sharpe, \textit{Differential Geometry: Cartan's Generalization of Klein's Erlangen Programme}}
	\item{Alexandrino and Bettiol \textit{Lie Groups and Geometric Aspects of Isometric Actions}}
	\item[]{Vassiliou, \textit{Geometry of Principal Sheaves}}
	\item[]{Morgan and Fong, \textit{Ricci Flow and Geometrization of 3-Manifolds} (aspirational)}
	\item[]{Jost, \textit{Riemannian Geometry and Geometric Analysis}}
	\item[]{Chow, Lu and Ni, \textit{Hamilton's Ricci Flow}}
\end{itemize}

\end{document}